### 2014-05-28 writeups/context-dependence.tex
.
intro/lit review

Interacting particle systems with neighborhood structure: transition probabilities generally not expressable (except TASEP).
Markov random fields with Glauber dynamics a case of this.

Example: Ising model.

Context-dependent substitution processes are an example of interest for phylogenetics and evolutionary inference.
Usually dealt with either by using larger and larger windows (but, only up to four bp),
or by doing MCMC over additional information (entire history of changes).
Also, Markov-along-the-genome method.

The difficulty comes from the fact that the rates of change at any one site depend in principle on arbitrarily far away sites,
so transition matrix is ginormous.

Goal: parameter inference, given before-after observations, or else observations descended from a common ancestor.

Review stats lit, e.g. reviews by Friel, original by Besag (autologistic model), genetics by Larribe and Fearnead.

Here restrict to one-dimensional neighborhood structure,
to make it easier.

\section{Notation and Examples}

@@

Since we cannot compute reasonably the full likelihood for the data,
we will need to choose appropriate summary statistics
that contain as much information about the process as possible,
and have a tractable expression for their marginal likelihood.
Assume here that $\ell$ has been chosen to be sufficiently large that approximation \eqref{eqn:window_approx} holds exactly.
If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
-- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ --
then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
are multinomial, with probabilities $\{p_{\ell,m}(x,y)\st y \in \calS^m\}$.
For instance, if $\ell=3$ then $N(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  C  & A & C & T & C & A & C \\
 $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
\end{tabular}
\end{center}
where $\cdot$ represents any base.
More generally, if we partition the sequence into windows of length $2\ell+m$,
and let $N(x,y)$ be the number of times we see a window in which $x$ matches $X(0)$ and $y$ matches $X(t)$,
then these counts are independently multinomial across each $x$.
% so the marginal likelihood of the statistics $N(x,y)$ is
% \begin{align}
%     \prod_{x \in \calS^{2\ell+m}} N(x)! \prod_{y \in \calS^m} \frac{p_{\ell,m}(x,y)^{N(x,y)}}{N(x,y)!} ,
% \end{align}
% where $N(x) = \sum_y N(x,y)$.

>>>>

It not uncommon to encounter models 
consisting of a lattice of sites,
each of which taking one of a finite collection of possible states,
and whose stochastic, temporal evolution is Markov and governed by a set of local rules.
For instance, at each site may sit 
a nucelotide whose mutation rate depends on physical properties determined by the nearby DNA sequence;
a cell whose infection status depends on the state of its neighbors;
or a particle whose spin is perturbed by external noise to a state whose distribution depends on the local energy configuration.
Practical use of such models often requires
inferring transition rules 
based on observations of the system at several time points,
or of several states evolved along a tree from a single starting point.
In this paper,
we propose one solution to this problem,
motivated by the problem of inferring context-dependent mutation rates from diverged nucleotide sequences.

The general class of models we consider are known in the probability literature as
interacting particle systems \citep{liggett2005ips} with neighborhood structure
-- continuous-time Markov chains on lattice-indexed collections of states
whose transition probabilities are \emph{local}
in the sense that any instantaneous change only affects a small number of nearby sites,
and the rates of such instantaneous changes depend only on the states in some bounded neighborhood of those sites to change.
As finite-state Markov chains,
transition probabilities are in principle simply expressable as a matrix exponential,
but this is impractical as the size of the matrix is equal to the number of possible configurations.
As an alternative, we show that the conditional likelihood of local pattern count statistics
can be well-approximated by marginalizing over a finite amount of surrounding context.

The situation is similar to the widely-studied problem of inference based on observation 
of a single instance of a Markov random field,
and in many cases reduces to this if we only observe the system at one time point at stationarity.
These are used, for instance, in spatial statistics \citep{besag1972nearestneighbour,gelfand2010handbook}
and image reconstruction \citep{geman1984stochastic,besag1986statistical}.
The conditioning method we consider here
is simliar to the ``coding'' scheme introduced by \citet{besag1974spatial},
that conditions on a set of sites that makes the remaining observations independent (thanks to the Gibbs property).
A good review of recent statistical techniques is given by \citet{friel2012bayesian}.

In the context of genomics,
it is well-known that certain short nucleotide sequences 
are in many organisms much more, or less, abundant than expected by chance \citep{karlin},
in part due to context dependence of the nucleotide mutation and substitution processes.
Phylogenetic and population genetic methods usually ignore such dependencies
in the interest of computational efficiency and lack of understanding of the underlying process,
but some progress has been made.
XXX
Usually dealt with either by using larger and larger windows (but, only up to four bp),
or by doing MCMC over additional information (entire history of changes).
Also, Markov-along-the-genome method.
Larribe and Fearnhead.

Another canonical example is the Ising model of statistical physics
with time evolution given by Glauber dynamics \citep{glauber1963timedependent},
in which a lattice of up/down spins are perturbed by thermal noise, 
relaxing into states dependent on the energy of the resulting configuration.
Parameter estimation for Ising model without temporal dynamics
is relatively well-undersood \citep{pickard1982inference,frigessi1990parameter},
but there does not seem to be similar work on the case when more than one observation is made of the dynamical system.

The general framework also fits certain cellular automata models,
e.g.\ modeling wildfire \citep{clarke1994cellular}, 
the spread of HIV \citep{zorzenondossantos2001dynamics}, 
or land use patterns \citep{clarke1996methods}.
Complex models may introduce long-range dependencies beyond the scope of this paper,
but these methods may still prove useful in the modeling process.

%%%%%%% %%%%%%%%%%
\section{Notation and Examples}

@@

Since we cannot compute reasonably the full likelihood for the data,
we will need to choose appropriate summary statistics
that contain as much information about the process as possible,
and have a tractable expression for their marginal likelihood.
For simplicity, 
we assume in this section that approximation \eqref{eqn:window_approx} holds exactly.
If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
-- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ --
then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
are multinomial, with probabilities $\{p_{\ell,m}(x,y)\st y \in \calS^m\}$.
For instance, if $\ell=3$ then $N(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  C  & A & C & T & C & A & C \\
 $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
\end{tabular}
\end{center}
where $\cdot$ represents any base.
More generally, if we partition the sequence into windows of length $2\ell+m$,
and let $N(x,y)$ be the number of times we see a window in which $x$ matches $X(0)$ and $y$ matches $X(t)$,
then these counts are independently multinomial across each $x$.
% so the marginal likelihood of the statistics $N(x,y)$ is
% \begin{align}
%     \prod_{x \in \calS^{2\ell+m}} N(x)! \prod_{y \in \calS^m} \frac{p_{\ell,m}(x,y)^{N(x,y)}}{N(x,y)!} ,
% \end{align}
% where $N(x) = \sum_y N(x,y)$.
